---
title: "Daily Weather Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daily Weather Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

Run this setup first. It loads the package, `dplyr`, `ggplot2`, `tidyr`, and
`cowplot`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.alt = "ggplot2 chart showing daily weather aggregation."
)
```

```{r packages, message=FALSE, warning=FALSE}
library(windcut)
library(dplyr)
library(ggplot2)
library(tidyr)
library(cowplot)
```

# Why aggregate to daily data?

Window-pane analysis can be done with hourly or daily data. For many plant
disease workflows, daily data are easier to inspect and easier to explain:
temperature can be represented by daily mean, minimum, or maximum; rainfall by
daily total; and leaf wetness by daily wet hours or daily proportion.

The important point is that the aggregation is an analysis decision. `windcut`
does not need to hide it. Use `aggregate_weather_daily()` when your source data
are hourly or sub-daily and you want to define the daily values before creating
candidate windows.

## Start from hourly weather

The bundled window-pane demo contains both daily weather and the hourly source
weather. The main workflow tutorials use the daily table, but this tutorial
starts from `weather_hourly` to show how that table can be created.

```{r}
data(window_pane_demo_data)

hourly_weather <- window_pane_demo_data$weather_hourly

hourly_weather %>%
  filter(site_id == site_id[1]) %>%
  slice_head(n = 8) %>%
  knitr::kable()
```

## Use the default daily summaries

The default daily aggregation keeps the usual `windcut` weather-column names.
Temperature and relative humidity are averaged, rainfall is summed, and
leaf-wetness observations are summed. With hourly source data, the daily
leaf-wetness sum can be interpreted as wet hours. The output column names always
show the time scale, statistic, and variable, such as `daily_mean_temp` and
`daily_sum_rain`.

```{r}
daily_weather <- aggregate_weather_daily(
  weather = hourly_weather,
  id_col = "site_id"
)

daily_weather %>%
  filter(site_id == site_id[1]) %>%
  slice_head(n = 8) %>%
  knitr::kable()
```

The result has one row per site per day. The `time` column is placed at
midnight so the daily table can be used directly with `make_windows()` and
`window_pane(unit = "days")`.

```{r}
daily_weather %>%
  count(site_id, name = "n_days") %>%
  knitr::kable()
```

## Choose the time and date columns

Many datasets have a timestamp column such as `time`, `datetime`, or
`timestamp`. If the data do not already have a date column,
`aggregate_weather_daily()` derives the day from `time_col` and writes it to
`date_col`.

```{r}
timestamp_weather <- hourly_weather %>%
  rename(timestamp = time)

daily_from_timestamp <- aggregate_weather_daily(
  weather = timestamp_weather,
  id_col = "site_id",
  time_col = "timestamp",
  date_col = "weather_date"
)

daily_from_timestamp %>%
  filter(site_id == site_id[1]) %>%
  select(site_id, weather_date, timestamp, daily_mean_temp, daily_mean_rh, daily_sum_rain) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

Some field datasets already have both a timestamp column and a day column. In
that case, set `date_col` to the existing day column. The function will use that
column for grouping and still create a daily timestamp column if
`keep_time = TRUE`.

```{r}
weather_with_day <- hourly_weather %>%
  mutate(weather_day = as.Date(time))

daily_from_existing_day <- aggregate_weather_daily(
  weather = weather_with_day,
  id_col = "site_id",
  time_col = "time",
  date_col = "weather_day"
)

daily_from_existing_day %>%
  filter(site_id == site_id[1]) %>%
  select(site_id, weather_day, time, daily_mean_temp, daily_mean_rh, daily_sum_rain) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

If the input is already daily and only has a date column, use
`time_col = NULL` and `keep_time = FALSE`. This is useful when you want to
redefine daily statistics from repeated daily records or harmonize an already
daily table without creating a timestamp column.

```{r}
already_dated <- hourly_weather %>%
  mutate(weather_day = as.Date(time)) %>%
  select(site_id, weather_day, temp, rain)

daily_from_date_only <- aggregate_weather_daily(
  weather = already_dated,
  id_col = "site_id",
  time_col = NULL,
  date_col = "weather_day",
  weather_cols = c("temp", "rain"),
  statistics = list(temp = "mean", rain = "sum"),
  keep_time = FALSE
)

daily_from_date_only %>%
  filter(site_id == site_id[1]) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

## Choose different daily statistics

The `statistics` argument uses the same idea as the window functions. A
character vector applies the same statistics to every selected weather
variable. This is compact when the same summaries make sense for all variables.

```{r}
daily_many_stats <- aggregate_weather_daily(
  weather = hourly_weather,
  id_col = "site_id",
  statistics = c("mean", "min", "max")
)

daily_many_stats %>%
  filter(site_id == site_id[1]) %>%
  select(site_id, date, time, daily_mean_temp, daily_min_temp, daily_max_temp, daily_mean_rh, daily_min_rh, daily_max_rh) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

A named list is better when each weather variable needs its own daily
definition. In this example, temperature gets mean, minimum, and maximum;
relative humidity gets mean and median; rainfall gets total and maximum hourly
rainfall; and leaf wetness gets wet hours.

```{r}
daily_selected_stats <- aggregate_weather_daily(
  weather = hourly_weather,
  id_col = "site_id",
  statistics = list(
    temp = c("mean", "min", "max"),
    rh = c("mean", "median"),
    rain = c("sum", "max"),
    leaf_wetness = list(wet_hours = "sum")
  )
)

daily_selected_stats %>%
  filter(site_id == site_id[1]) %>%
  select(site_id, date, starts_with("daily_mean_temp"), starts_with("daily_sum_rain"), daily_wet_hours_leaf_wetness) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

## Use non-standard weather-column names

Field datasets often use names such as `temp2m`, `prectot`, or `sradiation`.
Use `weather_cols` to tell `windcut` which columns should be aggregated. The
simplest approach is to pass the real column names and use those same names in
`statistics`.

```{r}
custom_hourly <- hourly_weather %>%
  rename(
    temp2m = temp,
    relhum = rh,
    prectot = rain
  ) %>%
  mutate(sradiation = pmax(0, 600 * sin(as.numeric(format(time, "%H")) / 24 * pi)))

custom_daily <- aggregate_weather_daily(
  weather = custom_hourly,
  id_col = "site_id",
  weather_cols = c("temp2m", "relhum", "prectot", "sradiation"),
  statistics = list(
    temp2m = c("mean", "max"),
    relhum = "mean",
    prectot = "sum",
    sradiation = "sum"
  )
)

custom_daily %>%
  filter(site_id == site_id[1]) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

If shorter output names are useful later, use a named `weather_cols` vector.
The names on the left become the names used in `statistics` and in the output
columns; the names on the right are the columns in the original dataset.

```{r}
custom_daily_short_names <- aggregate_weather_daily(
  weather = custom_hourly,
  id_col = "site_id",
  weather_cols = c(
    air_temp = "temp2m",
    humidity = "relhum",
    rain = "prectot",
    solar = "sradiation"
  ),
  statistics = list(
    air_temp = c("mean", "max"),
    humidity = "mean",
    rain = "sum",
    solar = "sum"
  )
)

custom_daily_short_names %>%
  filter(site_id == site_id[1]) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

## Check the daily signal

After aggregation, plot the daily weather before moving to window generation.
This is a simple way to catch unit mistakes, unexpected missing periods, or
daily statistics that do not match the biology of the disease.

```{r}
plot_daily <- daily_weather %>%
  filter(site_id == site_id[1]) %>%
  select(site_id, date, daily_mean_temp, daily_mean_rh, daily_sum_rain, daily_sum_leaf_wetness) %>%
  pivot_longer(
    cols = c(daily_mean_temp, daily_mean_rh, daily_sum_rain, daily_sum_leaf_wetness),
    names_to = "variable",
    values_to = "value"
  )

ggplot(plot_daily, aes(date, value)) +
  geom_line(color = "#3f7d58", linewidth = 0.7) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  labs(
    title = "Daily weather after aggregation",
    x = NULL,
    y = NULL
  ) +
  cowplot::theme_half_open()
```

## Use the daily data in the window-pane workflow

The daily table can be passed directly to `window_pane()`. Negative offsets
create windows before the reference date, positive offsets create windows after
the reference date, and mixed offsets create windows around the reference date.

```{r}
windows <- make_windows(
  min_offset = -21,
  max_offset = -1,
  width = 5,
  reference_col = "assessment_time"
)

features <- window_pane(
  weather = daily_weather,
  assessments = window_pane_demo_data$assessments,
  windows = windows,
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = c(
    temp = "daily_mean_temp",
    rh = "daily_mean_rh",
    rain = "daily_sum_rain",
    leaf_wetness = "daily_sum_leaf_wetness"
  ),
  unit = "days"
)

features %>%
  select(1:10) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

This separation keeps the workflow explicit: first define what one day means,
then define where the candidate windows sit relative to the biological
reference date.
