---
title: "Feature Screening for Disease Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature Screening for Disease Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

Run this setup first. It loads the packages used in the article and sets figure
options used by the screening examples.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.alt = "ggplot2 chart showing the result of this windcut workflow step."
)
```

```{r packages, message=FALSE, warning=FALSE}
library(windcut)
library(dplyr)
library(ggplot2)
library(tidyr)
library(cowplot)
```

# From windows to candidate predictors

Window-pane analysis can create many weather predictors. The next task is to
decide which ones deserve attention before modeling. This article uses two
screening functions:

- `screen_window_features()` ranks each candidate predictor by its correlation
  with the disease response.
- `screen_feature_correlations()` compares candidate predictors against each
  other and suggests a less redundant subset.

<div class="tutorial-note">
<strong>Important.</strong> Screening is not model validation. It is only an
first pass for reducing a large feature space before fitting predictive models.
</div>

## Build a feature table

Before screening, we need one row per site and one column per candidate
predictor. The bundled `window_pane_demo_data` object already contains daily
weather summaries and one disease assessment per site. Here the windows end one
day before the assessment date, so the disease observation is not mixed with
same-day weather.

The daily weather column names are kept exactly as they appear in the data:
`daily_mean_temp`, `daily_mean_rh`, `daily_sum_rain`, and
`daily_sum_leaf_wetness`. We also derive `daily_vpd` from temperature and
relative humidity to show that derived variables can be screened in the same
workflow. The `statistics` list uses those column names, so the generated
feature names show both the weather variable and the window summary.

```{r}
data(window_pane_demo_data)

weather <- window_pane_demo_data$weather |>
  derive_vpd(daily_mean_temp, daily_mean_rh, name = "daily_vpd")
assessments <- window_pane_demo_data$assessments

windows <- make_windows(min_offset = -18, max_offset = -1, width = 5, reference_col = "assessment_time")

daily_weather_cols <- c(
  "daily_mean_temp",
  "daily_mean_rh",
  "daily_vpd",
  "daily_sum_rain",
  "daily_sum_leaf_wetness"
)

summary_statistics <- list(
  daily_mean_temp = list(mean = "mean", days_18_26 = count_between(18, 26)),
  daily_mean_rh = list(mean = "mean", humid_days = humid_hours(90)),
  daily_vpd = list(mean = "mean", dry_days = count_above(1.2)),
  daily_sum_rain = list(total = "sum", rain_events = rain_events(0.2)),
  daily_sum_leaf_wetness = list(wet_days = wet_days(0)),
  .conditions = list(
    favorable_days = count_when(
      daily_mean_temp >= 18 & daily_mean_temp <= 26 & daily_mean_rh >= 90
    )
  )
)

features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = windows,
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

feature_cols <- features %>%
  select(contains("_window_")) %>%
  names()

feature_table_overview <- data.frame(
  rows = nrow(features),
  candidate_feature_columns = length(feature_cols)
)

knitr::kable(feature_table_overview)
knitr::kable(data.frame(feature = head(feature_cols, 8)))
```

## Screen with Spearman correlation

The first question is: which individual weather-window predictors are most
associated with disease intensity? `screen_window_features()` answers that with
one correlation test per candidate predictor. Spearman correlation is useful for
this example because it ranks monotonic relationships and is less dependent on a
linear response shape than Pearson correlation.

```{r}
screened <- screen_window_features(
  data = features,
  response_col = "disease_intensity",
  feature_cols = feature_cols,
  method = "spearman"
)

knitr::kable(screened %>% slice_head(n = 10))
```

Plotting the strongest screened features makes the sign and magnitude of the
associations easier to scan. Positive and negative correlations are separated
by color because they suggest different biological interpretations: higher
values of a predictor may be associated with either higher or lower disease
intensity.

```{r}
top10 <- screened %>%
  mutate(abs_estimate = abs(estimate)) %>%
  arrange(desc(abs_estimate)) %>%
  slice_head(n = 10) %>%
  mutate(feature = factor(feature, levels = rev(feature)))

ggplot(top10, aes(estimate, feature, fill = estimate > 0)) +
  geom_col(show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("#c47f2c", "#5f8f63")) +
  labs(
    title = "Top feature-screening results",
    x = "Correlation estimate",
    y = NULL
  ) +
  cowplot::theme_half_open()
```

The output provides:

- `estimate`: correlation estimate
- `p_value`: raw test p-value
- `p_adjusted`: corrected p-value
- `metric`: parsed metric name
- `window`: parsed relative-time window label

The adjusted p-value is useful because many window-derived predictors are being
tested at once. It should not be read as final proof that a predictor belongs in
a model, but it helps keep the first screen from being driven only by chance.

## Reduce correlated candidate features

Many modeling approaches become unstable or difficult to interpret when the
predictor set contains strongly correlated variables. Use
`screen_feature_correlations()` to compare candidate features against each other
and suggest a less redundant subset.

Here the correlation threshold is set to `0.8`, meaning pairs with absolute
correlation of at least 0.8 are treated as strongly redundant. The method is
user-controlled, so Pearson, Spearman, or Kendall can be chosen for the
problem. When a highly correlated pair is found, the function keeps the feature
that is less correlated with the rest of the candidate set.

```{r}
correlation_screen <- screen_feature_correlations(
  data = features,
  feature_cols = feature_cols,
  method = "spearman",
  threshold = 0.8
)

knitr::kable(correlation_screen$high_correlations %>% slice_head(n = 10))
knitr::kable(
  data.frame(suggested_feature = correlation_screen$suggested_features) %>%
    slice_head(n = 12)
)
```

The screening object also makes it easy to quantify how much redundancy was
removed. The table and plot below compare the original number of candidate
features with the number suggested for modeling and the number set aside under
the selected correlation threshold.

```{r}
feature_reduction <- data.frame(
  status = c("Original features", "Suggested features", "Removed features"),
  count = c(
    length(feature_cols),
    length(correlation_screen$suggested_features),
    length(correlation_screen$removed_features)
  )
)

knitr::kable(feature_reduction)

ggplot(feature_reduction, aes(count, status, fill = status)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = count), hjust = -0.25, size = 3.8) +
  scale_fill_manual(values = c(
    "Original features" = "#3f7d58",
    "Suggested features" = "#6ea87d",
    "Removed features" = "#c47f2c"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Feature count after correlation screening",
    x = "Number of feature columns",
    y = NULL
  ) +
  cowplot::theme_half_open()
```

The first plot explains the suggestion mechanism. We plot the most redundant
variables first; features with high mean absolute correlation to the rest of the
feature set are more likely to be set aside. Green bars are suggested for the
next modeling step; orange bars are redundant under the selected threshold.

```{r, fig.height = 9}
summary_plot <- correlation_screen$feature_summary %>%
  arrange(desc(mean_abs_correlation)) %>%
  slice_head(n = 30) %>%
  mutate(feature = factor(feature, levels = rev(feature)))

ggplot(summary_plot, aes(mean_abs_correlation, feature, fill = suggested)) +
  geom_col() +
  scale_fill_manual(values = c("#c47f2c", "#5f8f63")) +
  labs(
    title = "Feature redundancy ranking",
    x = "Mean absolute correlation with other features",
    y = NULL,
    fill = "Suggested"
  ) +
  cowplot::theme_half_open()
```

The heatmap complements the ranking plot by showing the pairwise structure.
Values close to 1 or -1 indicate features carrying very similar information,
while values close to 0 indicate features that are less redundant. To keep
labels readable, this plot uses a focused subset of the suggested features.

```{r, fig.width = 8, fig.height = 8}
heatmap_features <- head(correlation_screen$suggested_features, 12)
heatmap_matrix <- correlation_screen$correlation_matrix[
  heatmap_features,
  heatmap_features,
  drop = FALSE
]

heatmap_data <- as.data.frame(as.table(heatmap_matrix)) %>%
  rename(feature_x = Var1, feature_y = Var2, correlation = Freq) %>%
  mutate(
    label = sprintf("%.2f", correlation),
    feature_x = factor(feature_x, levels = heatmap_features),
    feature_y = factor(feature_y, levels = rev(heatmap_features))
  )

ggplot(heatmap_data, aes(feature_x, feature_y, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = label), size = 2.6) +
  scale_fill_gradient2(
    low = "#4b79a8",
    mid = "white",
    high = "#b13a32",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation"
  ) +
  labs(
    title = "Correlation matrix for suggested features",
    x = NULL,
    y = NULL
  ) +
  coord_fixed() +
  cowplot::theme_half_open() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid = element_blank()
  )
```

## A practical interpretation workflow

1. Look for metrics that show repeated signal across nearby windows.
2. Prefer windows that make biological sense, not only statistical sense.
3. Remove or set aside highly correlated predictors when the modeling approach
   assumes low feature redundancy.
4. Avoid overinterpreting isolated peaks when neighboring windows are inconsistent.
5. Carry a small candidate set into the predictive modeling stage.




