---
title: "Functional Data Analysis for Weather Series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional Data Analysis for Weather Series}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

Run this setup first. It loads `windcut`, the small tidyverse pieces used for
displaying tables, and `cowplot` for the plot theme. It also keeps chunk
messages out of the rendered tutorial.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8,
  dev.args = list(bg = "white"),
  fig.alt = "ggplot2 chart from the functional data analysis workflow."
)
```

```{r packages, message=FALSE, warning=FALSE}
library(windcut)
library(dplyr)
library(ggplot2)
library(cowplot)
```

# Why FDA in `windcut`?

Window-pane analysis stratifies a time series into user-defined windows and
computes summary statistics inside each one. Functional data analysis asks a
different question: whether the full weather trajectory differs between disease
classes and, if so, where along the time axis that difference appears.

This vignette is inspired by the functional data analysis and feature
engineering strategy described by [Alves et al. (2025)](https://doi.org/10.1094/PHYTO-01-25-0008-FI).
The package workflow is more general than the analysis in that paper: the user
chooses the weather variables, smoothing method, p-value threshold, and number
of permutations.

The main user-facing function is `run_fda_analysis()`. It does the repetitive
work that users should not have to code by hand: building subject-by-time
matrices, computing smoothed group means, running interval-wise tests for every
weather variable, and storing tidy outputs for plots and feature extraction.

The lower-level functions are still available for advanced users, but most FDA
workflows should start with this sequence:

| Step | Function |
|:--|:--|
| Run the workflow | `run_fda_analysis()` |
| Plot class mean curves | `plot_fda_means()` |
| Plot class differences | `plot_fda_difference()` |
| Plot corrected p-values | `plot_fda_p_values()` |
| Plot significant intervals | `plot_fda_intervals()` |
| Extract modeling features | `extract_fda_features()` |

<div class="tutorial-note">
<strong>Important.</strong> This vignette uses the bundled `fda_demo_data`
dataset so the FDA workflow stays focused on analysis and interpretation.
</div>

## Step 1: Load the demo data

The demo object contains daily weather series, one binary disease assessment per
site, and display labels for the weather variables. The disease classes are
`wm-` and `wm+`, matching the binary white mold example.

```{r}
data(fda_demo_data)

weather_daily <- fda_demo_data$weather_daily
assessments <- fda_demo_data$assessments
variable_specs <- fda_demo_data$variable_specs

knitr::kable(head(weather_daily))
knitr::kable(head(assessments))
knitr::kable(variable_specs)
```

The FDA functions expect the disease class to be available in the same long
table as the weather measurements. This is ordinary data preparation: one row is
one site at one time point.

```{r}
subject_data <- weather_daily %>%
  left_join(assessments %>% select(site_id, wm_class), by = "site_id")

knitr::kable(
  subject_data %>%
    select(site_id, dap, daily_mean_rh, daily_sum_rain, daily_mean_leaf_wetness, wm_class) %>%
    slice_head(n = 6)
)
```

Before running FDA, it is worth checking that both disease classes are present.

```{r}
class_counts <- assessments %>%
  count(wm_class, name = "n_sites")

knitr::kable(class_counts)
```

## Step 2: Run the FDA workflow

This single call is the core of the vignette. The user tells `windcut` which
columns identify the site, time axis, disease class, and weather variables. The
same call also chooses the smoothing method, p-value thresholds, and number of
permutations.

```{r}
fda_fit <- run_fda_analysis(
  data = subject_data,
  id_col = "site_id",
  time_col = "dap",
  group_col = "wm_class",
  value_cols = variable_specs$variable,
  value_labels = variable_specs,
  smooth_method = "spline",
  smooth_args = list(spar = 0.65),
  alpha = c(0.05, 0.01),
  n_permutations = 200
)
```

The object stores the decisions used in the analysis and tidy interval results.
The interval table already includes variables with no significant interval, so
the user does not need to write special-case code for them.

```{r}
knitr::kable(fda_fit$interval_summary)
```

## Step 3: Inspect one weather variable

Relative humidity is the clearest signal in this demo dataset. The mean-curve
plot shows the smoothed average trajectory for each disease class.

```{r}
plot_fda_means(
  fda_fit,
  value_col = "daily_mean_rh",
  xlab = "Days after planting",
  ylab = "Mean relative humidity (%)",
  legend_title = "White mold class",
  palette = c("wm-" = "#2b6c4f", "wm+" = "#c47f2c")
)
```

The difference plot shows the same comparison as `wm+` minus `wm-`. Positive
values mean the `wm+` sites had higher relative humidity on those days.

```{r}
plot_fda_difference(
  fda_fit,
  value_col = "daily_mean_rh",
  xlab = "Days after planting",
  ylab = "wm+ minus wm-"
)
```

The corrected p-value plot shows where the class difference is strongest. The
dashed lines are the alpha thresholds chosen in `run_fda_analysis()`.

```{r}
plot_fda_p_values(
  fda_fit,
  value_col = "daily_mean_rh",
  alpha = c(0.05, 0.01),
  xlab = "Days after planting"
)
```

## Step 4: Compare significant intervals across variables

The interval plot summarizes the interval-wise tests for all weather variables.
Each segment is a time period where the two disease classes differ for that
variable. Variables with no significant interval are labeled explicitly.

```{r}
plot_fda_intervals(
  fda_fit,
  title = "Significant FDA intervals by variable",
  xlab = "Days relative to planting"
)
```

Using two alpha values is useful because the user can see how conservative
thresholds change the candidate periods. In this example, `alpha = 0.05` is the
main feature-engineering threshold, while `alpha = 0.01` shows the stricter
alternative.

## Step 5: Fit a function-on-scalar model

The interval-wise test is the main feature-selection tool in this workflow. A
function-on-scalar model is a complementary view: it estimates the disease-class
effect across the whole time axis for one weather variable. The function below
uses the groups already stored in `fda_fit`, so the user does not need to build
a contrast vector manually.

```{r}
rh_model <- fit_fda_group_model(
  fda_fit,
  value_col = "daily_mean_rh",
  bs.yindex = list(bs = "ps", k = 20, m = c(2, 1)),
  bs.int = list(bs = "ps", k = 20, m = c(2, 1))
)
```

The coefficient plot shows the estimated disease-class effect over time. Values
farther from zero indicate time periods where relative humidity separates the
two disease classes more strongly.

```{r}
plot_functional_on_scalar(
  rh_model,
  title = "Function-on-scalar coefficient function for relative humidity",
  xlab = "Days after planting",
  ylab = "Coefficient function"
)
```

## Step 6: Extract features for modeling

The final step converts significant FDA intervals into ordinary site-level
predictors. The result is a regular data frame: one row per site and one column
per interval-derived feature. The `statistics` argument controls how each
significant interval is summarized for each site.

When the same summaries make sense for every variable, give `statistics` as a
character vector.

```{r}
fda_features <- extract_fda_features(
  fda_fit,
  alpha = 0.05,
  prefix = "fda",
  statistics = c("mean", "max")
)

knitr::kable(head(fda_features))
```

Different variables often need different summaries. For example, humidity may
be represented by the interval mean and minimum, rainfall by the interval sum
and maximum, and leaf wetness by the mean and total exposure. The `.default`
entry is optional; it is used for variables not named explicitly.

```{r}
fda_features_by_variable <- extract_fda_features(
  fda_fit,
  alpha = 0.05,
  prefix = "fda",
  statistics = list(
    daily_mean_temp = c("mean", "min", "max"),
    daily_mean_rh = c("mean", "min"),
    daily_sum_rain = list(total = "sum", max = "max"),
    daily_mean_leaf_wetness = list(mean = "mean", total = "sum"),
    .default = "mean"
  )
)

knitr::kable(head(fda_features_by_variable))
```

Named functions can also be used when the statistic should have a biological
interpretation beyond the usual mean, minimum, maximum, or sum.

```{r}
fda_features_biological <- extract_fda_features(
  fda_fit,
  alpha = 0.05,
  prefix = "fda",
  statistics = list(
    daily_mean_rh = list(mean = "mean", humid_days = humid_hours(85)),
    daily_mean_temp = list(mean = "mean", days_18_26 = count_between(18, 26)),
    daily_sum_rain = list(total = "sum", rainy_days = rainy_days(0))
  )
)

knitr::kable(head(fda_features_biological))
```

Those features can be joined to disease outcomes or other site-level covariates
for downstream modeling. The plot below uses ggplot2 directly because this is a
general modeling diagnostic, not part of the FDA algorithm itself.

```{r}
feature_name <- setdiff(names(fda_features), "site_id")[1]

feature_plot_data <- fda_features %>%
  left_join(assessments %>% select(site_id, wm_class), by = "site_id") %>%
  transmute(site_id, wm_class, value = .data[[feature_name]])

ggplot(feature_plot_data, aes(wm_class, value, fill = wm_class, color = wm_class)) +
  geom_violin(width = 0.85, alpha = 0.22, linewidth = 0.4, show.legend = FALSE) +
  geom_boxplot(width = 0.18, alpha = 0.6, outlier.shape = NA, linewidth = 0.5, show.legend = FALSE) +
  geom_jitter(width = 0.08, alpha = 0.55, size = 1.7, show.legend = FALSE) +
  scale_fill_manual(values = c("wm-" = "#2b6c4f", "wm+" = "#c47f2c")) +
  scale_color_manual(values = c("wm-" = "#2b6c4f", "wm+" = "#c47f2c")) +
  labs(
    title = "Interval-derived FDA feature by binary disease class",
    x = "White mold class",
    y = feature_name
  ) +
  cowplot::theme_half_open()
```

## What the workflow gives the user

The FDA workflow now keeps the repetitive coding inside the package. The user
mainly decides which biological comparison to make:

1. which site and time columns define the curves,
2. which disease classes are compared,
3. which weather variables are screened,
4. which smoothing method is used,
5. which p-value thresholds define candidate intervals.

The lower-level functions remain available for users who want to customize a
single step, but the standard workflow should require only a few package
functions.

## References

Alves, K. S., Shah, D. A., Dillard, H. R., Del Ponte, E. M., and Pethybridge,
S. J. 2025. Safer and smarter: Leveraging interpretation-guided modeling and
data merging of disease and environmental data for plant disease risk
prediction. *Phytopathology* 115(10):1329-1343.




