---
title: "Getting Started with windcut"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with windcut}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

Run this setup first. It loads the packages used in the tutorial and sets
figure options.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.alt = "ggplot2 chart showing simulated weather, weather windows, or screened features."
)
```

```{r packages, message=FALSE, warning=FALSE}
library(windcut)
library(dplyr)
library(ggplot2)
library(tidyr)
library(cowplot)
```

`windcut` is designed for plant disease epidemiologists who want to explore
which periods of weather are most informative around a biologically meaningful
reference date.

<div class="tutorial-note">
<strong>What you will learn.</strong> In this tutorial, we will load a bundled
demo dataset, define relative-time windows, generate weather features, and
screen them against disease intensity.
</div>

## Load a demo dataset

The first step is to load a bundled multi-site dataset. Each site has its own
daily weather time series, and the assessment table contains one disease
outcome per site. This mirrors the package's modeling unit: weather history for
a location paired with one disease outcome. The bundled data also include the
hourly source table used to create the daily version, but this tutorial starts
from daily data so the window-pane workflow is easier to read.

```{r}
data(window_pane_demo_data)

weather <- window_pane_demo_data$weather
assessments <- window_pane_demo_data$assessments

knitr::kable(head(weather))
knitr::kable(assessments)
```

Before generating features, it is useful to look at the daily weather signal
relative to a reference date. The dashed line marks the disease assessment for
one site, and the plotted weather history is the period immediately before that
reference date that later gets summarized into windows.

```{r}
first_reference_time <- assessments %>%
  slice(1) %>%
  pull(assessment_time)
plot_weather <- weather %>%
  filter(site_id == assessments$site_id[1]) %>%
  filter(time >= first_reference_time - 21 * 86400, time <= first_reference_time)

ggplot(plot_weather, aes(time, daily_mean_temp)) +
  geom_vline(
    xintercept = first_reference_time,
    linetype = "dashed",
    color = "#20262e",
    linewidth = 0.7
  ) +
  geom_line(color = "#2b6c4f", linewidth = 0.8) +
  labs(
    title = "Temperature history before the assessment",
    x = "Time",
    y = "Daily mean temperature (deg C)"
  ) +
  cowplot::theme_half_open()
```

## Understand the data structure

The bundled demo weather is already daily. Its weather-column names include the
daily statistic used to create them:

- `time`: timestamp of the daily observation
- `daily_mean_temp`: daily mean temperature
- `daily_mean_rh`: daily mean relative humidity
- `daily_sum_rain`: daily rainfall total
- `daily_sum_leaf_wetness`: daily wet hours

And an assessment table with at least:

- a reference date, such as assessment, planting, flowering, or inoculation date
- optionally an assessment ID
- optionally a disease response such as incidence or severity

Your own data do not need to use those exact weather-column names. In
`scan_windows()` and `window_pane()`, use `weather_cols` to choose the weather
variables to summarize. For example, a dataset with columns named `temp2m`,
`sradiation`, and `prectot` can use `weather_cols = c("temp2m", "sradiation",
"prectot")`. If you want shorter feature names, use a named vector such as
`weather_cols = c(temp = "temp2m", solar = "sradiation", rain = "prectot")`.

## Define candidate windows

Here we use the classic fixed-width window pane: one 5-day exposure period
slides through the relative-time range. Negative offsets are before the
reference date, `0` is the reference date, and positive offsets are after the
reference date. The optional `reference_col` stores which date column should be
used later when the windows are applied to site-specific data. The default
`slide_by = 1` moves the window one relative-time unit at a time; use a larger
value when you want a coarser scan.

```{r}
windows <- make_windows(
  min_offset = -21,
  max_offset = 0,
  width = 5,
  reference_col = "assessment_time"
)

knitr::kable(head(windows))
```

Window labels encode direction and interval boundaries. For example,
`window_m05_z00` means that a weather summary was computed over the 5-day
interval ending at whichever reference date the user chooses. A label such as
`window_p01_p06` would mean the 5-day interval beginning 1 day after the
reference date.

When the relevant exposure duration is uncertain, give `width` several values,
such as `width = 2:4` or `width = c(2, 4, 6)`. The same `slide_by` argument
controls how far the window start moves between candidate windows.

Because the demo weather uses daily column names, we define which daily columns
should be summarized inside each candidate window. The names on the left become
the names used in the generated window features.

```{r}
daily_weather_cols <- c(
  temp = "daily_mean_temp",
  rh = "daily_mean_rh",
  rain = "daily_sum_rain",
  leaf_wetness = "daily_sum_leaf_wetness"
)
```

The `statistics` argument controls what is computed inside each candidate
window. For example, `statistics = c("mean", "median", "sd", "IQR")` asks for
those four R functions for each weather variable. Custom named functions also
work.

## Scan one reference date

Scanning one site and one reference date first is a good diagnostic step. It shows
exactly which summaries will be computed for each candidate window before the
same operation is repeated for every site.

```{r}
scanned <- scan_windows(
  weather = weather %>% filter(site_id == assessments$site_id[1]),
  reference_time = assessments$assessment_time[1],
  windows = windows,
  weather_cols = daily_weather_cols,
  statistics = list(
    temp = c("mean", "median"),
    rh = list(mean = "mean", days_ge_90 = count_ge(90)),
    rain = c("sum", "sd"),
    leaf_wetness = "sum"
  )
)

knitr::kable(
  scanned %>%
    select(label, temp_mean, temp_median, rain_sum, rain_sd, rh_mean, rh_days_ge_90) %>%
    slice_head(n = 6)
)
```

This is useful when you want to inspect one assessment in detail before scaling
up to all observations.

## Build a model-ready table

`window_pane()` scales the single-assessment scan to the full dataset. The
result is a wide table with one row per site and one column for each
metric-window combination, ready for screening or modeling.

```{r}
features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = windows,
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = list(
    temp = c("mean", "median"),
    rh = list(mean = "mean", days_ge_90 = count_ge(90)),
    rain = c("sum", "sd"),
    leaf_wetness = "sum"
  )
)

feature_overview <- data.frame(
  rows = nrow(features),
  columns = ncol(features)
)
feature_names <- data.frame(
  feature = names(features %>% select(1:10))
)

knitr::kable(feature_overview)
knitr::kable(feature_names)
```

## Screen candidate features

The first screening question is whether each window-derived feature is related
to the disease response. Spearman correlation is used here because it can
capture monotonic relationships without assuming linearity.

```{r}
screened <- screen_window_features(
  data = features,
  response_col = "disease_intensity",
  method = "spearman"
)

knitr::kable(head(screened))
```

The table is easier to interpret when the strongest associations are plotted.
The chart below ranks the top features by absolute correlation, highlighting
which weather summaries and timing windows deserve closer inspection.

```{r}
top_screened <- screened %>%
  mutate(abs_estimate = abs(estimate)) %>%
  arrange(desc(abs_estimate)) %>%
  slice_head(n = 8) %>%
  mutate(feature = factor(feature, levels = rev(feature)))

ggplot(top_screened, aes(abs(estimate), feature)) +
  geom_col(fill = "#6ea87d") +
  labs(
    title = "Top screened windows",
    x = "Absolute correlation",
    y = NULL
  ) +
  cowplot::theme_half_open()
```

This output can be passed to downstream modeling workflows for feature
selection, disease forecasting, or explanatory analyses.

## Suggested workflow

1. Start with broad windows that reflect epidemiological uncertainty.
2. Generate interpretable summaries rather than opaque features.
3. Screen candidate predictors to identify promising timing intervals.
4. Fit and validate predictive models on independent datasets whenever possible.




