---
title: "Summarizing multivariable disease-favorable conditions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarizing multivariable disease-favorable conditions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

Many plant disease processes depend on combinations of weather variables.
Temperature alone may not be enough, and humidity alone may not be enough, but
warm and humid periods together can represent biologically favorable exposure.
The `.conditions` entry in `statistics` lets a user summarize those
multivariable rules inside each candidate window.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.alt = "ggplot2 chart showing multivariable disease-favorable weather conditions."
)
```

```{r packages, message=FALSE, warning=FALSE}
library(windcut)
library(dplyr)
library(ggplot2)
library(tidyr)
library(cowplot)
```

## Function glossary

These functions are used inside the special `.conditions` entry of
`statistics`. Each condition function receives the whole weather window, so the
expression can combine temperature, humidity, rain, wetness, or any other
columns in the weather table.

| Function | What it does | Typical use |
|---|---|---|
| `count_when()` | Count rows where a multivariable condition is `TRUE` | Number of favorable observations |
| `proportion_when()` | Return the fraction of valid rows where the condition is `TRUE` | Compare favorable exposure across windows |
| `sum_when()` | Sum one variable only when the condition is `TRUE` | Rainfall during warm periods |
| `mean_when()` | Average one variable only when the condition is `TRUE` | Mean humidity during warm periods |
| `max_when()` / `min_when()` | Return the maximum or minimum of one variable when the condition is `TRUE` | Extremes during biologically favorable periods |
| `max_consecutive_when()` | Find the longest uninterrupted run where the condition is `TRUE` | Longest favorable spell |
| `spell_count_when()` | Count separate uninterrupted runs where the condition is `TRUE` | Number of favorable episodes |
| `mean_spell_duration_when()` | Average the length of favorable spells | Typical spell duration |
| `max_spell_duration_when()` | Return the length of the longest favorable spell | Maximum continuous exposure |

## Load weather and disease data

The condition expressions refer to the actual column names in the weather
table. In this demo, the daily columns are named `daily_mean_temp`,
`daily_mean_rh`, `daily_sum_rain`, and `daily_sum_leaf_wetness`.

```{r}
data(window_pane_demo_data)

weather <- window_pane_demo_data$weather
assessments <- window_pane_demo_data$assessments

knitr::kable(head(weather))
```

## Make one condition visible

Before using `.conditions`, it helps to draw the rule on the time series. Here a
day is called favorable when temperature is between 18 and 26 degrees C and
relative humidity is at least 85%. The orange points are the observations that
satisfy the full rule. The threshold is chosen to make the demo visible; a real
analysis should use thresholds justified by the pathosystem.

```{r}
example_site <- assessments$site_id[1]
example_reference <- assessments %>%
  filter(site_id == example_site) %>%
  pull(assessment_time)

condition_weather <- weather %>%
  filter(site_id == example_site) %>%
  filter(time >= example_reference - 35 * 86400, time <= example_reference) %>%
  mutate(
    favorable = daily_mean_temp >= 18 &
      daily_mean_temp <= 26 &
      daily_mean_rh >= 85,
    wet_and_warm = daily_sum_leaf_wetness > 0 &
      daily_mean_temp >= 18 &
      daily_mean_temp <= 26
  )

ggplot(condition_weather, aes(time, daily_mean_temp)) +
  geom_vline(
    xintercept = example_reference,
    linetype = "dashed",
    color = "#20262e",
    linewidth = 0.7
  ) +
  geom_ribbon(aes(ymin = 18, ymax = 26), fill = "#6ea87d", alpha = 0.12) +
  geom_line(color = "#20262e", linewidth = 0.8) +
  geom_point(
    data = condition_weather %>% filter(favorable),
    color = "#c47f2c",
    size = 2.6
  ) +
  labs(
    title = "A multivariable condition is a biological rule applied row by row",
    subtitle = "Orange points satisfy temperature and humidity requirements",
    x = NULL,
    y = "Daily mean temperature (deg C)"
  ) +
  theme_half_open()
```

The same rule can be shown as a calendar-like condition timeline. This is often
the easiest plot for checking whether the condition creates isolated favorable
days or longer spells.

```{r}
condition_timeline <- condition_weather %>%
  select(time, favorable, wet_and_warm) %>%
  pivot_longer(
    cols = c(favorable, wet_and_warm),
    names_to = "condition",
    values_to = "present"
  )

ggplot(condition_timeline, aes(time, condition, fill = present)) +
  geom_tile(color = "white", linewidth = 0.5, height = 0.8) +
  scale_fill_manual(values = c("FALSE" = "#e8ecef", "TRUE" = "#2b6c4f")) +
  labs(
    title = "Condition timelines reveal duration and interruption",
    x = NULL,
    y = NULL,
    fill = "Condition met"
  ) +
  theme_half_open() +
  theme(panel.grid = element_blank())
```

## Define multivariable summaries

A condition summary receives the whole weather window as a data frame. That is
why expressions can combine several variables. Output names from `.conditions`
are not prefixed by one weather-variable name because the condition already
describes a multivariable process.

```{r}
summary_statistics <- list(
  daily_mean_temp = list(
    mean = "mean",
    days_18_26 = count_between(18, 26)
  ),
  daily_mean_rh = list(
    mean = "mean",
    humid_days = humid_hours(85)
  ),
  daily_sum_rain = list(
    total = "sum"
  ),
  daily_sum_leaf_wetness = list(
    wet_days = wet_days(0)
  ),
  .conditions = list(
    favorable_days = count_when(
      daily_mean_temp >= 18 & daily_mean_temp <= 26 & daily_mean_rh >= 85
    ),
    favorable_spell = max_consecutive_when(
      daily_sum_leaf_wetness > 0 & daily_mean_temp >= 18 & daily_mean_temp <= 26
    ),
    warm_rain_total = sum_when(
      daily_sum_rain,
      daily_mean_temp >= 18 & daily_mean_temp <= 28
    )
  )
)
```

## Apply the rules inside candidate windows

The window-pane workflow does not change. The only difference is that
`statistics` now includes `.conditions`.

```{r}
windows <- make_windows(
  min_offset = -35,
  max_offset = 0,
  width = 7,
  slide_by = 7,
  reference_col = "assessment_time"
)

features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = windows,
  reference_col = "assessment_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  unit = "days",
  statistics = summary_statistics
)

features %>%
  select(site_id, disease_intensity, contains("favorable"), contains("warm_rain")) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

## Compare condition functions in the output

`favorable_days` counts how many rows inside a window satisfy the full
temperature-humidity rule. `favorable_spell` measures the longest uninterrupted
wet-and-warm period. `warm_rain_total` sums rain only on rows where temperature
is inside the selected range.

This article stops at interpretation of the generated variables. It does not try
to demonstrate a disease-response relationship, because that belongs in a later
screening or modeling workflow.

```{r}
condition_examples <- features |>
  select(site_id, contains("favorable"), contains("warm_rain")) |>
  slice_head(n = 6)

knitr::kable(condition_examples)
```

The plot below uses one site to show how the condition summaries differ across
candidate windows. The bars are not a model result; they are a check that each
condition function produced the intended biological quantity.

```{r}
condition_features <- features |>
  filter(site_id == example_site) |>
  select(contains("favorable_days"), contains("favorable_spell"), contains("warm_rain_total")) |>
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "value"
  ) |>
  mutate(
    window = sub(".*_window_", "window_", feature),
    summary_class = case_when(
      grepl("favorable_days", feature) ~ "Count of favorable days",
      grepl("favorable_spell", feature) ~ "Longest favorable spell",
      grepl("warm_rain_total", feature) ~ "Rain during warm days",
      TRUE ~ "Condition summary"
    ),
    window = factor(window, levels = unique(window))
  )

ggplot(condition_features, aes(window, value, fill = summary_class)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ summary_class, scales = "free_y", ncol = 1) +
  labs(
    title = "Condition functions summarize different biological quantities",
    subtitle = "Example values for one site across 7-day candidate windows",
    x = "Candidate window",
    y = "Summary value"
  ) +
  theme_half_open() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Use `.conditions` when the biological rule depends on more than one weather
variable. Use variable-specific summaries when the statistic only needs one
column, such as mean temperature, total rain, or humid observations.
