---
title: "Window-Pane Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Window-Pane Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

Run this setup first. It loads the packages used in the article and sets figure
options.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8,
  dev.args = list(bg = "white"),
  fig.alt = "ggplot2 chart showing reference dates, candidate windows, or generated weather features."
)
```

```{r packages, message=FALSE, warning=FALSE}
library(windcut)
library(dplyr)
library(ggplot2)
library(tidyr)
library(cowplot)
```

# Why window-pane?

Plant disease symptoms are often influenced by weather near a biological event.
The event may be planting, flowering, inoculation, disease assessment, or
another reference date. The hard part is that the most informative period is
usually unknown. A window-pane workflow solves this by generating many candidate
relative-time windows, summarizing weather inside each one, and comparing the
resulting features.

<div class="tutorial-note">
<strong>Main idea.</strong> Choose a reference date, decide whether candidate
windows should be before, after, or around that date, and summarize weather
inside each candidate window.
</div>

The reference date does not have to be the disease assessment date. It can also
be planting, flowering, inoculation, harvest, or any timestamp that makes sense
for the biology of the system.

## Step 1: Load bundled site-level weather and disease data

This tutorial starts from a bundled demo dataset with 10 sites. Each site has
one daily weather time series and one disease assessment. That structure
mirrors the modeling unit used by `windcut`: one row of disease data is paired
with the weather history from the same site.

```{r}
data(window_pane_demo_data)

weather <- window_pane_demo_data$weather
assessments <- window_pane_demo_data$assessments

nrow(weather)
knitr::kable(assessments)
```

The `weather` object is already daily. If your own data are hourly, use
`aggregate_weather_daily()` before this step and choose the daily statistics
that match each weather variable. Daily column names keep the statistic in the
name, such as `daily_mean_temp` and `daily_sum_rain`.

The `site_id` column is important because it lets `window_pane()` match each
assessment to the correct weather series. The `assessment_time` and
`planting_time` columns are both valid reference dates, but they answer
different questions.

## Step 2: Visualize the possible reference dates

Before defining windows, it helps to see what the reference dates mean on the
weather timeline. The plot below uses one site and draws two vertical reference
lines: planting and assessment. Later, the same relative-time grid will be placed relative
to either one of these columns by changing `reference_col`.

```{r}
example_site <- assessments$site_id[1]
example_assessment <- assessments %>% filter(site_id == example_site)
example_weather <- weather %>% filter(site_id == example_site)

plot_start <- example_assessment$planting_time - 7 * 86400
plot_end <- example_assessment$assessment_time + 1 * 86400
plot_weather <- example_weather %>%
  filter(time >= plot_start, time <= plot_end)

reference_dates <- data.frame(
  reference_col = c("planting_time", "assessment_time"),
  reference_time = c(
    example_assessment$planting_time,
    example_assessment$assessment_time
  )
)

ggplot(plot_weather, aes(time, daily_mean_temp)) +
  geom_line(color = "#3f7d58", linewidth = 0.7) +
  geom_vline(
    data = reference_dates,
    aes(xintercept = reference_time, color = reference_col),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  scale_color_manual(values = c(planting_time = "#c47f2c", assessment_time = "#20262e")) +
  labs(
    title = "Two possible reference dates for one site",
    x = "Calendar time",
    y = "Daily mean temperature (deg C)",
    color = "Reference column"
  ) +
  cowplot::theme_half_open()
```

In `make_windows()`, offsets are measured relative to the selected reference
date. Negative offsets are before the reference date, `0` is the reference date,
and positive offsets are after the reference date. Window labels describe
interval boundaries. For example, `window_m05_z00` means a 5-day interval that
ends at the selected reference date, while `window_p01_p06` means a 5-day
interval that begins 1 day after the reference date.

## Step 3: Create a fixed-width window pane

A fixed-width window pane uses one duration and slides it through the
relative-time range. Here every candidate window is 7 days long and occurs
before or on the selected reference date. To match the 49-day span used later
for planting-centered windows, this grid runs from 49 days before assessment to
the assessment date itself. The first window is 49 to 43 days before the
reference date, the next is 48 to 42 days before the reference date, then 47 to
41 days before the reference date, and so on. That default behavior is
controlled by `slide_by = 1`.

```{r}
fixed_windows <- make_windows(
  min_offset = -49,
  max_offset = 0,
  width = 7
)

knitr::kable(head(fixed_windows, 10))
nrow(fixed_windows)
```

When a dense one-day slide creates more candidate windows than needed, increase
`slide_by`. In the next grid, the same 7-day window moves two relative-time
units at a time: 49 to 43 days before the reference date, then 47 to 41 days
before the reference date, then 45 to 39 days before the reference date.

```{r}
coarser_windows <- make_windows(
  min_offset = -49,
  max_offset = 0,
  width = 7,
  slide_by = 2
)

knitr::kable(head(coarser_windows, 10))
nrow(coarser_windows)
```

The comparison below shows the practical effect of `slide_by`. The window width
is still 7 days in both grids, but a larger slide produces fewer, more widely
spaced candidate windows.

```{r}
slide_comparison <- data.frame(
  slide_by = c("slide_by = 1", "slide_by = 2"),
  candidate_windows = c(nrow(fixed_windows), nrow(coarser_windows))
)

knitr::kable(slide_comparison)
```

The first plot shows the dense fixed-width grid. The dashed line at
relative-time 0 is the reference date. Segments to the left of that line are before the
reference event.

```{r dense-fixed-window-plot, fig.height=9}
plot_window_pane(
  fixed_windows,
  max_windows = Inf,
  color_by = "none",
  title = "Dense fixed-width window-pane grid",
  subtitle = "slide_by = 1 moves the 7-day window one relative-time unit at a time",
  xlab = "Time relative to reference date (days)"
)
```

The next plot shows the same 7-day duration with a larger sliding step. The
grid is smaller because the starting relative-time position jumps by two units
instead of one.

```{r coarser-fixed-window-plot, fig.height=6.5}
plot_window_pane(
  coarser_windows,
  max_windows = Inf,
  color_by = "none",
  title = "Coarser fixed-width window-pane grid",
  subtitle = "slide_by = 2 moves the 7-day window two relative-time units at a time",
  xlab = "Time relative to reference date (days)"
)
```

Use fixed-width windows when the exposure duration is already biologically
defendable. For example, a 7-day period may be meaningful if infection requires
several consecutive days of favorable moisture and temperature.

## Step 4: Create a variable-width window pane

A variable-width window pane scans both timing and duration. Instead of using a
single duration, give `width` several values. Here the workflow tries 3-, 5-,
and 7-day windows within the same 49-day assessment-centered range. The window
starts move two relative-time units at a time because `slide_by = 2`.

```{r}
variable_windows <- make_windows(
  min_offset = -49,
  max_offset = 0,
  width = c(3, 5, 7),
  slide_by = 2
)

knitr::kable(head(variable_windows, 10))
nrow(variable_windows)
```

The variable-width plot uses color to show duration. Shorter and longer windows
can start at the same relative-time position, but they cover different portions of the weather
history before the same reference date. The gap between adjacent starting positions
shows the effect of `slide_by`.

```{r variable-window-plot, fig.height=12}
plot_window_pane(
  variable_windows,
  max_windows = Inf,
  color_by = "width",
  title = "Variable-width window-pane grid",
  subtitle = "The dashed line at relative-time 0 marks the reference date",
  xlab = "Time relative to reference date (days)"
)
```

The choice between fixed and variable windows should come from the disease
biology. Fixed windows are simpler and easier to interpret. Variable windows
are broader and may be better for discovery, especially early in a study.

## Step 5: Generate features using assessment dates

`window_pane()` applies a window grid to every site. For each assessment row, it
finds the matching weather series, places each candidate window relative to the
chosen `reference_col`, and returns one wide feature table. In this first call,
the windows are relative to `assessment_time`.

The `statistics` argument controls how weather is summarized inside each
candidate window. A named list is useful when different variables need different
summaries. In this example, the weather columns keep the names already present
in the daily table, such as `daily_mean_temp`, `daily_mean_rh`, and
`daily_sum_rain`. The same names are used in `statistics`, so the output feature
names make the full calculation explicit.

```{r}
daily_weather_cols <- c(
  "daily_mean_temp",
  "daily_mean_rh",
  "daily_sum_rain",
  "daily_sum_leaf_wetness"
)

summary_statistics <- list(
  daily_mean_temp = c("mean", "min", "max"),
  daily_mean_rh = list(mean = "mean", median = "median", days_at_or_above_90 = count_at_or_above(90)),
  daily_sum_rain = c("sum", "max"),
  daily_sum_leaf_wetness = "sum"
)

assessment_features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = fixed_windows,
  reference_col = "assessment_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

assessment_feature_overview <- data.frame(
  rows = nrow(assessment_features),
  columns = ncol(assessment_features)
)
assessment_feature_names <- data.frame(
  feature = names(assessment_features %>% select(1:12))
)

knitr::kable(assessment_feature_overview)
knitr::kable(assessment_feature_names)
```

The first columns identify the site, the reference date, and the response. The
remaining columns are weather summaries. For example,
`daily_mean_temp_mean_window_m07_z00` is the mean of `daily_mean_temp` over the
7-day interval ending at the assessment date.

The table below shows the first rows and first feature columns of
`assessment_features`. The full object has more weather-summary columns than
shown here, but this view is enough to see the shape of the output: one row per
site and one column per candidate weather predictor.

```{r}
assessment_features %>%
  select(1:12) %>%
  slice_head(n = 6) %>%
  knitr::kable()
```

## Step 6: Generate features using planting dates

Planting is positioned much earlier in the weather series than assessment, so
it deserves its own window grid. Around assessment, the usual question is often
"what happened before disease was measured?" Around planting, the question may
include weather before planting and weather after planting, because both can be
biologically meaningful.

The next grid uses 7-day windows and moves one week at a time. Negative windows
summarize weather before planting, the window that starts at `0` includes the
planting date, and positive windows summarize early-season weather after
planting.

```{r}
planting_windows <- make_windows(
  min_offset = -14,
  max_offset = 35,
  width = 7,
  slide_by = 7,
  reference_col = "planting_time"
)

knitr::kable(planting_windows)
```

The plot below makes the reference-date logic visible. The dashed line is
planting. Segments to the left are pre-planting windows; segments to the right
are post-planting windows.

```{r}
plot_window_pane(
  planting_windows,
  max_windows = Inf,
  color_by = "timing",
  title = "Planting-centered windows can cross both sides of the reference date",
  subtitle = "Negative offsets are before planting; positive offsets are after planting",
  xlab = "Time relative to planting date (days)"
)
```

Now `window_pane()` uses `planting_windows` with `reference_col =
"planting_time"`. The weather table and summary statistics stay the same; only
the biological clock and the candidate windows change.

```{r}
planting_features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = planting_windows,
  reference_col = "planting_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

knitr::kable(
  planting_features %>%
    select(site_id, planting_time, disease_intensity) %>%
    slice_head(n = 5)
)
```

This is useful when the relevant biological clock starts at planting rather
than assessment. For example, early-season weather may affect plant
establishment, host susceptibility, or the amount of inoculum available later.

## Step 7: Compare feature-table size across window strategies

Different window strategies create different numbers of candidate predictors.
For a fair comparison, the four grids below use the same 49-day
assessment-centered range introduced in Step 3: `-49` to `0`. This also matches
the total length of the planting-centered grid above, where `35 - (-14) = 49`.
The only differences are the window durations and the sliding step. This lets
the user see how a short sliding step creates a dense feature table, while a
longer sliding step creates a smaller, more widely spaced one.

```{r}
comparison_sparse_windows <- make_windows(
  min_offset = -49,
  max_offset = 0,
  width = 7,
  slide_by = 7
)

comparison_dense_features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = fixed_windows,
  reference_col = "assessment_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

comparison_coarser_features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = coarser_windows,
  reference_col = "assessment_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

comparison_variable_features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = variable_windows,
  reference_col = "assessment_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

comparison_sparse_features <- window_pane(
  weather = weather,
  assessments = assessments,
  windows = comparison_sparse_windows,
  reference_col = "assessment_time",
  id_col = "site_id",
  response_col = "disease_intensity",
  weather_cols = daily_weather_cols,
  statistics = summary_statistics
)

feature_table_sizes <- data.frame(
  window_strategy = c(
    "7-day windows, slide_by = 1",
    "7-day windows, slide_by = 2",
    "3-, 5-, and 7-day windows, slide_by = 2",
    "7-day windows, slide_by = 7"
  ),
  relative_range = rep("-49 to 0 days", 4),
  total_length_days = rep(49, 4),
  candidate_windows = c(
    nrow(fixed_windows),
    nrow(coarser_windows),
    nrow(variable_windows),
    nrow(comparison_sparse_windows)
  ),
  generated_features = c(
    sum(grepl("_window_", names(comparison_dense_features))),
    sum(grepl("_window_", names(comparison_coarser_features))),
    sum(grepl("_window_", names(comparison_variable_features))),
    sum(grepl("_window_", names(comparison_sparse_features)))
  )
)

knitr::kable(feature_table_sizes)
```

The plot below emphasizes the practical trade-off. The label above each bar is
the number of generated feature columns. Dense sliding and variable-width
windows search a larger feature space. A longer sliding step keeps the same
total relative-time range but creates fewer candidate predictors, which can be
useful when the goal is a smaller, more interpretable modeling table.

```{r feature-table-size-plot, fig.width=8, fig.height=4.8}
feature_table_sizes <- feature_table_sizes %>%
  mutate(window_strategy = factor(window_strategy, levels = window_strategy))

ggplot(feature_table_sizes, aes(generated_features, window_strategy, fill = window_strategy)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = generated_features), hjust = -0.25, size = 3.8) +
  scale_fill_manual(values = c(
    "7-day windows, slide_by = 1" = "#3f7d58",
    "7-day windows, slide_by = 2" = "#6ea87d",
    "3-, 5-, and 7-day windows, slide_by = 2" = "#c47f2c",
    "7-day windows, slide_by = 7" = "#7f5539"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Feature-table size by window strategy",
    subtitle = "Same 49-day range; different duration and slide_by",
    x = "Generated feature columns",
    y = NULL
  ) +
  cowplot::theme_half_open()
```

## Step 8: Inspect one metric family

After feature generation, inspect groups of related variables before modeling.
Here we focus on temperature means from the fixed-width, assessment-referenced
table. The object `temp_cols` contains only feature names that start with
`daily_mean_temp_mean_`.

```{r}
features <- assessment_features
temp_cols <- features %>%
  select(starts_with("daily_mean_temp_mean_")) %>%
  names()
knitr::kable(data.frame(feature = head(temp_cols)))
```

The plot below converts a small set of temperature features from wide to long
format using `tidyr::pivot_longer()`. Each line is a site, and each point is a candidate
temperature window. This helps reveal whether sites differ consistently across
windows or only in specific timing periods.

```{r}
plot_ids <- features %>%
  slice_head(n = min(4, nrow(features))) %>%
  pull(site_id)
plot_cols <- temp_cols[1:min(8, length(temp_cols))]
temp_long <- features %>%
  filter(site_id %in% plot_ids) %>%
  select(site_id, all_of(plot_cols)) %>%
  pivot_longer(
    cols = -site_id,
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(feature_index = match(feature, plot_cols))

ggplot(temp_long, aes(feature_index, value, color = site_id, group = site_id)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c("#2b6c4f", "#8aa05a", "#c47f2c", "#7f5539")) +
  scale_x_continuous(
    breaks = seq_along(plot_cols),
    labels = sub("daily_mean_temp_mean_", "", plot_cols)
  ) +
  labs(
    title = "Temperature summaries across candidate windows",
    x = "Relative-time window",
    y = "Feature value",
    color = "Site"
  ) +
  cowplot::theme_half_open() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The metric family often matters as much as the window timing:

- temperature may relate to infection efficiency
- leaf wetness may relate to germination and penetration
- rainfall may relate to dispersal or splash
- relative humidity may act as a proxy for microclimatic favorability

## Step 9: What to do next

Window-pane analysis can generate many predictors. A careful workflow usually
screens candidates, removes redundant variables, and then fits models using a
small set of biologically defendable features. Use `screen_window_features()` to
rank features by association with disease intensity, and use
`screen_feature_correlations()` to identify highly correlated predictors before
modeling.




